---
title: "project"
format: html
editor: visual
---

```{r}
data <- read_csv("../../CAINC4_MN_1969_2022.csv")   #importing income dataset
demographic_all <- read.csv("../../cc-est2019-alldata-27.csv")   #importing demographic dataset
head(demographic_all)

demographic_all %>% group_by(COUNTY, CTYNAME) %>% summarize(count = n())
```

## Data viz

### variable units for "income_year" data

-   personal income (thousands of dollars)
-   per capita personal income (dollars)
-   wages and salaries (Thousands of dollars)
-   non-farm personal income (thousands of dollars)
-   farm income (thousands of dollars)
-   total employment (Number of jobs)
-   population (Number of persons)

```{r}

# Cleaning the income dataset using Ellie's code, and leaving variables 
filtered_data <- data[data$Description %in% c("Personal income (thousands of dollars)", "Per capita personal income (dollars) 4/", "Wages and salaries", "Nonfarm personal income 1/", "Farm income 2/", "Total employment", "Population (persons) 3/"), ]
filtered_data <- filtered_data[, !(names(filtered_data) %in% c("Region", "LineCode", "TableName", "Unit", "IndustryClassification"))]
filtered_data <- na.omit(filtered_data) %>% select(-(`1969`:`2011`))

# adjusting income with inflation rates using Ellie's code!
#This is where the interest rate change is
cpi_data <- data.frame(
  Year = 2012:2022,
  CPI = c(229.6, 233.0, 236.7, 237.0, 240.0, 245.1, 251.1, 255.7, 258.8, 271.0, 292.7) 
)

# Set base year for CPI adjustment (2022)
base_year <- 2022
cpi_base <- cpi_data %>% filter(Year == base_year) %>% pull(CPI)

# Adjust each year from 2012 to 2022 based on CPI
adjusted_data <- filtered_data %>%
  mutate(across(starts_with("20"), 
                ~ . * cpi_base / cpi_data %>% filter(Year == as.numeric(substring(cur_column(), 1, 4))) %>% pull(CPI),
                .names = "Real_{col}"))

income_year <- adjusted_data %>% 
  select(-(`2012`:`2022`)) %>% 
  rename("2012" = "Real_2012", "2013" = "Real_2013", "2014" = "Real_2014", "2015" = "Real_2015", "2016" = "Real_2016", "2017" = "Real_2017", "2018" = "Real_2018", "2019" = "Real_2019", "2020" = "Real_2020", "2021" = "Real_2021", "2022" = "Real_2022")

income_year <- income_year %>% 
  pivot_longer(`2012`:`2022`,names_to = 'year',values_to='value') %>%
  pivot_wider(names_from = Description, values_from = value)

income_year <- income_year %>% 
  rename(per_capita_income = 'Per capita personal income (dollars) 4/',
         personal_income = 'Personal income (thousands of dollars)',
         salaries = 'Wages and salaries',
         population = 'Population (persons) 3/',
         total_employment = 'Total employment',
         non_farm_income = 'Nonfarm personal income 1/',
         farm_income = 'Farm income 2/')

income_year <- income_year %>% 
  rename(county = "GeoName")

income_year <- income_year %>% 
  mutate(county = str_remove_all(county, ", MN"))

income_year %>% head()

```

```{r}
# Combining male and female demographics
demographic_all <- demographic_all %>%
  mutate(White = WA_MALE + WA_FEMALE,
         Black = BA_MALE + BA_FEMALE,
         Hispanic = H_MALE + H_FEMALE,
         Indian_Alaska = IA_MALE + IA_FEMALE,
         Asian = AA_MALE + AA_FEMALE,
         Hawaiian_PI = NA_MALE + NA_FEMALE,
         Two_or_more = TOM_MALE + TOM_FEMALE)

# selecting important variables
demographic_all <- demographic_all %>% 
  select(STATE, COUNTY, CTYNAME, YEAR, TOT_POP, TOT_MALE, TOT_FEMALE, AGEGRP, White, Black, Hispanic, Indian_Alaska, Asian, Hawaiian_PI, Two_or_more)

# Formating the years and county name
demographic_data <- demographic_all %>% 
  filter(AGEGRP == 0) %>% 
  filter(YEAR != 1 & YEAR != 2) %>% 
  mutate(YEAR = YEAR + 2007) %>%
  mutate(CTYNAME = str_remove(CTYNAME, " County")) 

# Creating a column combining non-white demographics
demographic_data <- demographic_data %>% 
  group_by(CTYNAME, YEAR) %>% 
  mutate(non_white = sum(Hispanic, Black, Indian_Alaska, Asian, Hawaiian_PI, Two_or_more)) %>%   select(-AGEGRP, -STATE)

```

```{r}
# Pivoting longer?
mn_demographic_long <- demographic_all %>%
  select(YEAR, CTYNAME, White, Black, Hispanic, Indian_Alaska, Asian, Hawaiian_PI, Two_or_more) %>%
  pivot_longer(cols = c(White, Black, Hispanic, Indian_Alaska, Asian, Hawaiian_PI, Two_or_more),
               names_to = "Race",
               values_to = "Population")



mn_demographic_long %>% arrange(desc(YEAR)) %>% head()
```

##demographic data plot
```{r}
# Different plots showing different distributions in different counties filtered
demographic_data %>% filter(CTYNAME == "Ramsey") %>% na.omit() %>% 
  ggplot(aes(x = non_white)) +
  geom_segment(aes(y = White, xend = c(tail(non_white, n = -1), NA), 
                   yend = c(tail(White, n = -1), NA)),
               arrow = arrow(length = unit(0.4, "cm")),
               color = 4) +
  geom_point(aes(y = White), size = 2, color = 4) +
  geom_text(aes(label = YEAR, x = non_white + non_white*0.02, y = White)) +
  theme_light()
  
```

##demographic data plot
```{r}
demographic_data %>% filter(CTYNAME == "Aitkin") %>% na.omit() %>% 
  ggplot(aes(x = non_white, y = White)) +
  geom_line(size = 1, color = 4) +
  geom_point(size = 2.5, color = 4) +
  geom_text(aes(label = YEAR, hjust = 1)) +
  theme_light()
```

##function to create the percent change
```{r}
year_percentage_change <- function(x){
    (x - x[1]) / x[1] * 100}
```

##failed income plot
```{r}

income_year %>% 
  filter(county == "Aitkin") %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = non_farm_income, color = "non_farm"))
```

## plot of different income changes
```{r}
income_change <- income_year %>% 
  mutate(`2012` = Real_2012 - Real_2012,
         `2013` = (Real_2013 - Real_2012)/Real_2012 * 100,
         `2014` = (Real_2014 - Real_2012)/Real_2012 * 100,
         `2015` = (Real_2015 - Real_2012)/Real_2012 * 100,
         `2016` = (Real_2016 - Real_2012)/Real_2012 * 100,
         `2017` = (Real_2017 - Real_2012)/Real_2012 * 100,
         `2018` = (Real_2018 - Real_2012)/Real_2012 * 100,
         `2019` = (Real_2019 - Real_2012)/Real_2012 * 100,
         `2020` = (Real_2020 - Real_2012)/Real_2012 * 100,
         `2021` = (Real_2021 - Real_2012)/Real_2012 * 100,
         `2022` = (Real_2022 - Real_2012)/Real_2012 * 100)

income_change <- income_change %>% 
  select(-GeoFIPS, -starts_with("Real_"))
```


```{r}
# install.packages("devtools")
# devtools::install_github("ricardo-bion/ggradar")
# library(ggradar)
income_change_new <- income_change %>% filter(GeoName=="Aitkin, MN") %>% select(-GeoName)

ggradar(income_change_new, 
        grid.min = -190,
        grid.max = 40) 
 

```














